# R Examples

## How to Install

To install the package from [CRAN](https://CRAN.R-project.org/package=deadwood), call:

```r
install.packages("deadwood")
```

::::{note}
For best performance, advanced users will benefit from compiling the package
from sources:

```r
Sys.setenv(CXX_DEFS="-O3 -march=native")  # for gcc and clang
install.packages(c("quitefastmst", "deadwood"), type="source")
```

To enable OpenMP on macOS, execute `brew install libomp`, and then call in R:

```r
Sys.setenv(CXX_DEFS="-O3 -march=native -Xpreprocessor -fopenmp")
install.packages(c("quitefastmst", "deadwood"), type="source")
```
::::


## Basic Use

An example noisy dataset[^datasetsource]:

[^datasetsource]: The discussed dataset comes from
G. Karypis, E.H. Han, V. Kumar,
CHAMELEON: A hierarchical clustering algorithm using dynamic modeling,
*IEEE Transactions on Computers* **32**(8), 68-75, 1999
and is available for download from
[the clustering benchmarks repository](https://clustering-benchmarks.gagolewski.com/).

```{r r_chameleon_t7_10k_dataset,fig.cap="The chameleon_t7_10k dataset",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
plot_scatter <- function(X, labels=1L) {
    if (is.logical(labels)) labels <- 1L-labels
    labels[is.na(labels)] <- 0L
    pal <- c("#aaaaaa55", palette.colors(palette="R4", alpha=0.3))
    plot(X, asp=1, ann=FALSE, col=pal[labels+1L], pch=16)
}

X1 <- as.matrix(read.table("chameleon_t7_10k.data.gz"))
plot_scatter(X1)
```


Let us perform outlier detection with *Deadwood*.

```{r r_chameleon_t7_10k_deadwood,fig.cap="Outlier detection in chameleon_t7_10k",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
library("deadwood")
is_outlier <- deadwood(X1, max_k=1)
plot_scatter(X1, is_outlier)
```


Fraction of detected outliers:

```{r}
mean(is_outlier)
```



## Clusters of Unequal Densities

The above dataset consists of clusters of relatively equal densities.
Here is an example featuring non-homogeneous subgroups.

```{r r_chameleon_t8_8k_dataset,fig.cap="Outlier detection in the chameleon_t8_8k dataset, without and with automatic subcluster detection",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
X2 <- as.matrix(read.table("chameleon_t8_8k.data.gz"))

par(mfrow=c(1, 2))
is_outlier1 <- deadwood(X2, max_k=1)
plot_scatter(X2, is_outlier1)

is_outlier2 <- deadwood(X2)
plot_scatter(X2, is_outlier2)
```

In the second case, Deadwood was able to identify three subclusters automatically. In each subcluster, the outlierness threshold was estimated independently.

```{r r_chameleon_t8_8k_dataset2,fig.cap="Detected subclusters in chameleon_t8_8k dataset",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
plot_scatter(X2, attr(is_outlier2, "cluster"))
```


## Clusters of Highly Imbalanced Sizes

The above dataset consists of relatively large clusters.
[wut/z2](https://clustering-benchmarks.gagolewski.com/) is an
example where two small clusters on the right are treated as anomalous:

```{r r_z2_dataset,fig.cap="Outlier detection in the z2 dataset",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
X3 <- as.matrix(read.table("z2.data.gz"))
is_outlier <- deadwood(X3)
plot_scatter(X3, is_outlier)
```

To remedy this, we can decrease the appropriate size threshold
(`max_debris_size`) parameter:

```{r r_z2_dataset2,fig.cap="Outlier detection in z2 with smaller max_debris_size",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
is_outlier <- deadwood(X3, max_debris_size=20)
plot_scatter(X3, is_outlier)
```

Alternatively, we can first identify the clusters in the dataset
(currently, [*Lumbermark*](https://lumbermark.gagolewski.com/)
and [*Genie*](https://genieclust.gagolewski.com/) are supported).
Then, we can perform outlier detection in each cluster separately:

```{r r_z2_lumbermark,fig.cap="Detected clusters and outliers in z2",echo=-1}
par(mar=c(2, 2, 0.5, 0.5))
library("lumbermark")
clusters <- lumbermark(X3, 5)
is_outlier <- deadwood(clusters)
plot_scatter(X3, attr(is_outlier, "cluster"))
```


::::{note}
*To learn more about R, check out my open-access textbook*
[Deep R Programming](https://deepr.gagolewski.com/).
::::
